---
title: "Normality Tester Function"
author: "Edgar Derricho"
date: "2025-05-14"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Observation of Mixture Distribution

Let's observe a mixture distribution of two normal samples with different standard deviations.
```{r}
# -- Create Data --
set.seed(123)
x1 <- rnorm(500, mean = 0, sd = 1)
x2 <- rnorm(500, mean = 0, sd = 3)
newDist <- c(x1, x2)

# -- Plot Data and Density and QQ Plot --
hist(newDist, probability = TRUE, col = "lightblue", main = "Mixture Histogram")
curve(dnorm(x, mean = mean(newDist), sd = sd(newDist)), col = "red", lwd = 2, add = TRUE)

# Q-Q Plot
qqnorm(newDist, main = "Q-Q Plot of Mixture")
qqline(newDist, col = "red", lwd = 2)

```

This data passes a 'bell curve' visual test and a qq-plot visual test, though there is some discrepancy in the tails.

Let's observe more a more robust means of testing the data:

Shapiro-Wilk Test
```{r}
shapiro.test(newDist)
```

Jrque-Bera Test
```{r}
library(tseries)

jarque.bera.test(newDist)
```

# Tester Function

```{r}
library(nortest)
library(tseries)

normalityTester <- function(data){
  # -- Libraries --
  library(nortest)
  library(tseries)
  library(e1071)
  
  # -- Minor cleaning --
  data <- as.numeric(na.omit(data)) # Omit NA Values and Create Numeric
  
  
  # -- Tests --
  sw <- shapiro.test(data)
  ad <- nortest::ad.test(data)
  ks <- ks.test(data, "pnorm")
  jb <- tseries::jarque.bera.test(data)
  
  # -- Store Results --
  shapiroRes <- ifelse(sw$p.value >= 0.05,
                       "Normal",
                       "Not Normal") # Shapiro Wilk
  andersonRes <- ifelse(ad$p.value >= 0.05,
                       "Normal",
                       "Not Normal") # Anderson-Darling
  kolmogorvRes <- ifelse(ks$p.value >= 0.05,
                       "Normal",
                       "Not Normal")
  jarqueRes <- ifelse(jb$p.value >= 0.05,
                       "Normal",
                       "Not Normal")
  skewnessRes <- ifelse(skewness(data) > -0.5 & skewness(data < 0.5),
                        "Within Bounds",
                        "Outside of Bounds")
  kurtosisRes <- ifelse(skewness(data) > -2 & skewness(data < 2),
                        "Within Bounds",
                        "Outside of Bounds")

  # -- Print Results --
  cat(
      "      Normality Tester Function         ", "\n", 
      "               ----------             ","\n",
      "     Shapiro-Wilk Test: ", shapiroRes, "\n",
      "     Anderson-Darling Test: ", andersonRes, "\n",
      "     Kolmogrov-Smirnov Test: ", kolmogorvRes, "\n",
      "     Jarque-Bera Test: ", jarqueRes, "\n",
      "     Skewness: ", skewnessRes, "\n",
      "     Kurtosis: ", kurtosisRes, "\n",
      "               ----------             ", "\n", "\n"
      )
  
  
  print(ifelse(shapiroRes == "Normal" &
           andersonRes == "Normal" &
           kolmogorvRes == "Normal" &
           jarqueRes == "Normal" &
           kurtosisRes == "Within Bounds" &
           skewnessRes == "Within Bounds",
         "Data Appears to be Normal",
         "Data is not Normal"))
  
  # -- Visual Tests --
  # Histogram
  hist(data, main = "Histogram of Data", col = "lightblue")
  
  # Boxplot
  boxplot(data, main = "Boxplot of Data", col = "lightblue")
  
  # Q-Q Plot
  qqnorm(data, main = "Q-Q Plot of Data")
  qqline(data, col = "red", lwd = 2)
  
  # -- Data Frame --
  result <- list(
    Shapiro_Wilk_PValue = sw$p.value,
    Anderson_Darling_PValue = ad$p.value,
    Kolmogrov_Smirnov_PValue =  ks$p.value,
    Jarque_Bera_PValue = jb$p.value,
    Skewness = skewness(data),
    Kurtosis = kurtosis(data)
  )
  
  return(result)
}

test <- normalityTester(rnorm(500))
```

Through testing our function with rnorm(500), we see that each test shows the data are normal and the skewness and kurtosis are withing bounds. Let's observe the mixture from above.

```{r}
test <- normalityTester(newDist)
test
```

